{fairmetrics}: An R package for fairness evaluation metrics with confidence intervals

### 2025-05-29 paper.Rmd
Update paper.Rmd

<!--- 
Rewritten by Jesse
--->
Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for certain groups, particularly those defined by protected attributes such as race, gender, or age. Evaluating fairness is a critical aspect of model development, as biased models can perpetuate or exacerbate existing social inequalities.  The {fairmetrics} R package offers a user-friendly framework for rigorously evaluating numerous group-based fairness criteria, including metrics based on independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity). These criteria assess whether a model is equally accurate or well-calibrated across predefined groups so that appropriate bias mitigation strategies can be implemented. {fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions, and includes example an dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].

@@

Machine learning (ML) offers significant potential for predictive modelling in biomedical research [@rajpurkarAIHealthMedicine2022]. Despite its promise, there is substantial evidence that, without appropriate forethought and planning, ML models can introduce or exacerbate health inequities by making less accurate decisions for certain groups or individuals [@grote2022enabling]. While existing software can compute fairness metrics, none provide out-of-the-box statistical inference, leaving practitioners without guidance on the uncertainty around those metrics. As ML becomes increasingly embedded in healthcare systems, ensuring equitable model performance across diverse populations is essential[@Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024]. The {fairmetrics} R package fills this gap by offering a suite of popular group-fairness metrics along with bootstrap-based confidence intervals, enabling more rigorous and interpretable assessments of fairness in biomedical ML.

@@

Group fairness criteria are typically classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022]. The {fairmetrics} package computes a range of group fairness metrics together with bootstrap-based confidence intervals for uncertainty quantification.  The metrics implemented in the package are briefly described below.

@@

-   **Equal Opportunity:**  Compares disparities in false negative rates between two groups, quantifying any difference in missed positive cases.

@@

-   **Positive Class Balance:**  Compares the distribution of predicted probabilities among individuals whose true outcome is positive between groups, ensuring that the model does not favor one group over another in its positive predictions.

-   **Negative Class Balance:** Compares the distribution of predicted probabilities among individuals whose true outcome is negative between groups, ensuring that the model does not favor one group over another in its negative predictions.

@@

-   **Brier Score Parity:** Compares the Brier score—the mean squared error of probabilistic predictions—is similar across groups, indicating comparable calibration.

-   **Accuracy Parity:** Compares the overall accuracy of a predictive model is equivalent across different groups.

@@

<!--
1. Ben - The package doesn't require data splitting, some people may want to use the package and incorporate CV within training.  Also, sometimes people evaluate fairness metrics on validation data. So suggest you simplify text + figure. Just start with the input.  Also, it doesn't seem like multiple should be a separate node?  doesnt it just include all the 4 boxes (sep, suff, ind, other)? can this be depicted? Also suggest you make the names more descriptive "Separation-based Metrics", "Independence-based Metrics", etc.
--->
The primary input to the {fairmetrics} package is a data frame or tibble which containing the model's predictions, true outcomes, and the protected attribute in question. \hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}. It is possible to evaluate a model for a specific or multiple group fairness metrics.

![Workflow for using {fairmetrics} to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow.png)

<!---
Edited by Jesse
--->
A simple example of how to use the {fairmetrics} package is shown below. The example makes use of the `mimic_preprocessed` dataset, a pre-processed version of the the Indwelling Arterial Catheter (IAC) Clinical Dataset, from MIMIC-II clinical database^[The raw version of this data is made available by PhysioNet [@goldberger2000physiobank] and can be accessed in {fairmetrics} package by loading the `mimic` dataset.] [@raffa2016clinical; @raffa2016data]. This dataset consists of 1776 hemodynamically stable patients with resperatory failure, and includes demographic information (patient age and gender), vital signs, laboratory results, whether an IAC was used, and a binary outcome indicating wheter the patient died within 28 days of admission.

While the choice of fairness metric used is context dependent, we show all metrics available with the `get_fairness_metrics()` function. In this example, we evaluate the model's fairness with respect to the protected attribute `gender`. For conditional statistical parity, we condition  being `>=60` years old. The model is trained on a subset of the data, and predictions are made on a test set. 
<!---
1. Ben - need to add detail on what the detail is in the above paragraph. What does the data contaon, where did it come from, what is the sample size, etc.

@@

2. Ben - suggest you show more of the functions in the package - say that, while the choice of metric is context dependent, we show all metrics to show the full range of the package. Also include interpretation of the various results.  Simple way to do this is show, the multiple output then show an example of one of the separation-based criterion or other to show users they can specify what they want.  
Ben Response: Please see below for the example and review.
---->

>>>>

<!--- 
Rewritten by Jesse
--->
Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for certain groups, particularly those defined by protected attributes such as race, gender, or age. Evaluating fairness is a critical aspect of model development, as biased models can perpetuate or exacerbate existing social inequalities.  The {fairmetrics} R package offers a user-friendly framework for rigorously evaluating numerous group-based fairness criteria, including metrics based on independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity). These criteria assess whether a model is equally accurate or well-calibrated across predefined groups so that appropriate bias mitigation strategies can be implemented. {fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions, and includes example a dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].

@@

Machine Learning (ML) has been increasingly integrated into high-stakes domains that significantly impact individuals and society, including criminal justice, healthcare, finance, employment, and education. As ML systems become a commonplace in decision-making processes, mounting evidence suggest that they often exhibit bias towards gender and ethnicity. For example, in criminal justice systems, COMPAS, a widely used risk assessment tool in the US, incorrectly classified Black defendants as high-risk at nearly twice the rate of White defendants. As COMPAS was used in court to inform judicial decision-making, this bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to White defendants with similar risk profiles. Similarly, a healthcare algorithm deployed throughout the US to identify patients with complex health needs has shown significantly lower accuracy for Black patients relative to White patients. This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to White patients. Therefore, there is a urgent need for companies, engineers, and researchers to ensure that ML models are fair and equitable across different groups before they are deployed in real-world applications.

While existing software can compute fairness metrics, they provide only point estimates without quantifying the uncertainty around those metrics. This limitation prevents users from determining whether observed disparities between groups are statistically significant or merely the result of random variation due to finite sample size, potentially leading to incorrect conclusions about fairness violations. The {fairmetrics} R package addresses this gap by providing bootstrap-based confidence intervals (CIs) for both difference-based and ratio-based group fairness metrics, enabling users to make statistically grounded decisions about the fairness of their models.

# Scope

The `{fairmetrics}` package is designed to evaluate group fairness in the setting of binary classification with a binary protected attribute. This restriction reflects standard practice in the fairness literature and is motivated by several considerations. First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative. Second, group fairness is the most widely used framework for binary classification tasks. Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness. This focus enables {fairmetrics} to provide statistically grounded uncertainty quantification for group fairness metrics commonly applied in binary classification tasks across diverse application domains.

@@

Group fairness criteria are typically classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022]. Independence requires that the model’s predictions be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups. Separation requires independence between the prediction and the protected attribute conditional on the true outcome, so that the probability of a positive prediction is equal across protected groups within positive (or negative) outcome class. Sufficiency requires independence between the outcome and the protected attribute conditional on the prediction, implying that once the model’s prediction is known, the protected attribute provides no additional information about the true outcome.

@@

-   **Equal Opportunity:**  Compares disparities in false negative rates between groups, quantifying any difference in missed positive cases.

@@

<!---
Updated the two defs below. It only checks the first-moment of the predicted probabilities, does not check the distribution.

Also updated the name, Balance of Positive/Negative Class is the name the original paper used.
--->

-   **Balance for Positive Class:**  Compares if the average of predicted probabilities among individuals whose true outcome is positive is similar between groups.

-   **Balance for Negative Class:** Compares if the average of predicted probabilities among individuals whose true outcome is negative is similar between groups.

@@

-   **Brier Score Parity:** Compares the Brier score—the mean squared error of predicted probabilities—is similar across groups, indicating comparable calibration.

-   **Accuracy Parity:** Compares the overall accuracy of a predictive model is equivalent across groups.

@@

<!--
1. Ben - The package doesn't require data splitting, some people may want to use the package and incorporate CV within training.  Also, sometimes people evaluate fairness metrics on validation data. So suggest you simplify text + figure. Just start with the input.  Also, it doesn't seem like multiple should be a separate node?  doesnt it just include all the 4 boxes (sep, suff, ind, other)? can this be depicted? Also suggest you make the names more descriptive "Separation-based Metrics", "Independence-based Metrics", etc.

Ben Response: Created 3 different figures to review. I think v3 is the best. Please critique if there are any issues.
--->
The primary input to the {fairmetrics} package is a data frame or tibble which containing the model's predictions, true outcomes, and the protected attribute in question. \hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}. It is possible to evaluate a model for a specific or multiple group fairness metrics.

![Workflow for using {fairmetrics} to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow-v3.png)

<!---
Previously edited by Jesse, now edited by Ben
--->
A simple example of how to use the {fairmetrics} package is shown below. The example makes use of the `mimic_preprocessed` dataset, a pre-processed version of the the Indwelling Arterial Catheter (IAC) Clinical Dataset, from MIMIC-II clinical database^[The raw version of this data is made available by PhysioNet [@goldberger2000physiobank] and can be accessed in {fairmetrics} package by loading the `mimic` dataset.] [@raffa2016clinical; @raffa2016data]. This dataset consists of 1776 hemodynamically stable patients with respiratory failure, and includes demographic information (patient age and gender), vital signs, laboratory results, whether an IAC was used, and a binary outcome indicating whether the patient died within 28 days of admission.

While the choice of fairness metric used is context dependent, we show all metrics available with the `get_fairness_metrics()` function. In this example, we evaluate the model's fairness with respect to the protected attribute `gender`. For conditional statistical parity, we condition on patients older than 60 years old. The model is trained on a subset of the data, and predictions are made on a test set. 
<!---
1. Ben - need to add detail on what the detail is in the above paragraph. What does the data contaon, where did it come from, what is the sample size, etc.

@@

2. Ben - suggest you show more of the functions in the package - say that, while the choice of metric is context dependent, we show all metrics to show the full range of the package. Also include interpretation of the various results.  Simple way to do this is show, the multiple output then show an example of one of the separation-based criterion or other to show users they can specify what they want.

Ben Response: Please see below for the examples and review.
---->
